---
title: "Project 1 - Measuring reefs"
author: "Elizabeth Madin"
date: "`r format(Sys.Date())`"
output: 
  github_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(sciplot)
library(vegan)
library(plotrix)
library(plyr)
```

## Summary: 

This sub-module focuses on measuring and comparing coral patch reefs in the field and with satellite, aerial (airplane and drone) imagery of various spatial resolutions, as well as GPD tracking. 

We're using this exercize as a means to understand the general implications of satellite imagery spatial resolution on measuring features of interest. However, it has real-world applications in terms of coral reef area measurement (e.g., [Madin and Madin (2015) Conservation Biology](https://drive.google.com/file/d/1C_vg4Uju5aTMWh7-Pf0ZhgrRtXSfSMNK/view) and https://allencoralatlas.org)). 

## Goals

* Make connection between use of field data and remote (e.g., satellite; aerial) imagery to measure biological/habitat variables
* Highlight strengths and limitations of each approach
* Start thinking about ?s to ask/answer in your own research
* Provide coral reef field experience
* Contribute drone imagery to long-term reef monitoring dataset

## Components

* Devise/practice field measurement protocol (transect- and drone-based)
* Measure reefs in-situ
* Measure reefs w/satellite imagery of various spatial resolutions via:
    + Google Earth
    + Planet Explorer
* Compare patterns and measurement variability based on method, imagery spatial resolution, etc.
* Write up results as 'mini-paper'

## Materials

* transect tapes, weighted: 2 x 50m per group
* slates: 2
* metadata sheet (on underwater paper)
* mesh bag(s) to carry everything
* drone
* drone batteries + charger

## Protocols

We will focus on two patch reefs: 20 (small) and 22 (medium)

#### Part 1: Field measurements

Preparation:

* Brainstorm, plan, and practice measurements on land:
    + Perimeter only
* Establish protocol for field measurements
* Create datasheets on waterproof paper
* Gather remaining supplies for morning field trip

Measuring reefs:

* Measure each study reef based on methods in protocol
* Record data (including metadata - e.g., date/time, depth, etc.) on datasheet(s)
* Make sure not to leave any gear in the water!

Data entry:

* Open the .csv datasheet template (data/MeasuringReefs_template.csv) 
    + Do not add/delete columns since we will combine datasets later
    + Re-save file with your initials appended to the end of the filename (e.g., "MeasuringReefs_EPM.csv")
* Choose one person from each group to enter the group's field data
    + All others should delete the two rows (2 & 27) where column "method" = "field"
    + For the group's data enterer, enter field data into this datasheet
    + The column headings you’ll be filling in are:
        - perimeter_m
        - area_m2 (use the formula =((C2/(2*PI()))^2)*PI() to calculate this from perimeter)
        - date 
        - observer (as your two or three initials)
        - notes (if needed)

#### Part 2: Imagery measurements

Getting started:

* Open .kmz/.kml patch reef placemark file ("FieldSites.kmz") by dragging into Google Earth
* Open .csv datasheet you saved with your initials
    + Column headings you’ll be filling in are:
        - perimeter_m
        - area_m2
        - observer (as your two or three initials)
        - notes (if needed)

Measuring reefs:

* You will measure two of the patch reefs in Kane’ohe Bay for:
    + Perimeter
        - use the formula =((C2/(2*PI()))^2)*PI() to calculate area from perimeter) in .csv (you'll need to change cell C2 to whatever row it's in)
* Do one replicate of each reef (replication will come from pooling the group measurements)
* You will use both Google Earth and Planet Explorer to measure reefs using imagery spanning a range of spatial resolutions

1) **Google Earth**

    + Create a folder in My Places (Add -> Folder); give it a name that makes sense (eg, "KBay patch reefs")
        - Keep this folder highlighted from now on while you're measuring reefs so the measurements will go straight into that folder
    + Measure each of the study reefs for the following dates/spatial resolutions by using the Time Slider to select the following dates’ imagery:


        Date (yyyymmdd)   | Resolution (m)  | Satellite                             | Provider     | 
        :----------------- | :--------------- | :------------------------------------- | :------------ |
        20130115          | 0.15            | NA; aerial imagery                    | Unknown      |
        20130822          | 0.50            | WorldView-1/2, GeoEye-1, or Quickbird | DigitalGlobe |
        20110108          | ~1.00           | Ikonos                                | DigitalGlobe |
        20030412          | ~2.00           | EarlyBird-1 (?)                       | DigitalGlobe |


    + Use the polygon measurement tool to extract perimeter 
    + Manually copy measurements into .csv 
    + Save measurement: name as “PR## _ #.## _ ##” = PR[reef #] _ [resolution, in m] _ [your initials] (e.g., “PR20_0.15_EM”)
    + If any of your measurments show up in Temporary Places, drag saved measurements to the folder you created in My Places
    + Frequently save My Places! (File -> Save -> Save My Places)
    + _Note_: if you need to go back to measurements, right-click on outlines (or their corresponding saved names in My Places) -> Get Info 

2) **Planet Explorer**

    + Draw a rectangular AOI (area of interest) to cover Kane’ohe Bay
    + Visually compare the Google Earth placemarks with the Planet Explorer AOI to locate your study reefs 
    + Measure each of the study reefs for the following satellites/dates/spatial resolutions by using the left-hand options area to select the following imagery:

        Date (yyyymmdd)   | Resolution (m)  | Satellite                             | Provider     | 
        :----------------- | :--------------- | :------------------------------------- | :------------ |
        20190316          | ~3.00           | PlanetScope (formerly Dove CubeSats)  | Planet       |
        20160208          | 5.00            | RapidEye                              | Planet       |
        20180307          | ~10.00          | Sentinel-2                            | ESA          |
        20190328          | 30.00           | Landsat 8                             | USGS         |


    + To find this imagery most efficiently, set the following search parameters:
        - Make sure you’re in “Browse” mode 
        - Date range (right under search box) = 2016/01/01 - today
        - Drop down menu = Daily imagery
        - Cloud cover = 0-10%
        - Area coverage = 90-100%
        - Source = tick sources corresponding to “satellite” above 
        - _Tip_: do this satellite by satellite for faster searching
    + Use the “measure distance” tool to extract perimeter (hold shift for free hand measurement)
    + Manually copy measurements into .csv 
    + _Note_: there’s no way to save measurements, so double-check each entry for accuracy before moving on

## Part 3: Data analysis

Once all data is collected: 

* Check your dataset for any typos, missing values, etc.
* Create a folder called "data" within whatever folder you're using for this and put your dataset in it 
* Create a folder called "output" within whatever folder you're using for this (leave empty)
* Create a new project-specific repository in Github 
* Link your Github repo with your project directory (folder)
* Commit/push to the new project-specific repository you created in Github

If time permits for merging and analysis of group data:

* Send your repo's link to the rest of the group via our "remote-sensing-module" Slack channel
* Pull (download) everyone else’s datasets
* Merge others’ datasets with your dataset & save
* Commit/push revised (merged) dataset to your Github repo
* Once everyone’s data is merged, create a new R Markdown (.Rmd) file in R Studio
* Make some basic plots to help you explore any patterns in the data
* Run some statistical analyses to determine sources of variance, any significant trends, etc. 

Do some plots to look at patterns in the data (here are a few basic ones to get you started - you can copy this code or write your own):

_Note: you'll need to change the data source file name to match your own, and your results will be different than mine as a result._ 

```{r data import, echo=TRUE, results='hide', include=FALSE}
################################################
# import data, observe structure
################################################

data<-read.csv("data/MeasuringReefs_EM.csv", header = TRUE)

names(data)
head(data)
data$year<-substr(data$date,0,4) # add a column for year only
```


```{r data explore-1, echo=TRUE, results='hide', fig.height=5, fig.width=5, fig.align='center'}

################################################
# explore patterns
################################################

# Effect of method
boxplot(area_m2~method, data=data, na.ignore=TRUE, col="lightgrey",
     ylab="Patch reef area (m^2)", 
     xlab="Method", 
     main="Area versus method")
dev.copy(pdf,"output/method_area_effect.pdf")    # can change to .png, etc.; can change size to incr resolution
dev.off()

# Effect of imagery resolution
## on area
plot(area_m2~imagery_resolution_m, data=data, col="slategrey",
     xlim=c(0,10),
     xlab="Imagery spatial resolution (m)", 
     ylab="Patch reef area (m^2)", 
     main="Area versus imagery resolution")
res.area.lm=lm(area_m2 ~ imagery_resolution_m, data=data)
abline(res.area.lm, col = "black")
summary(res.area.lm)
dev.copy(pdf,"output/resolution_area_effect.pdf")    
dev.off()

## on perimeter
plot(perimeter_m~imagery_resolution_m, data=data, col="slategrey",
     xlim=c(0,10),
     xlab="Imagery spatial resolution (m)", 
     ylab="Patch reef perimeter (m)", 
     main="Perimeter versus imagery resolution")
res.perim.lm=lm(perimeter_m ~ imagery_resolution_m, data=data)
abline(res.perim.lm, col = "black")
summary(res.perim.lm)
dev.copy(pdf,"output/resolution_perimeter_effect.pdf")    
dev.off()

# Effect of date (i.e., do patch reefs measurably grow or shrink over time?)
plot(area_m2~as.numeric(year), data=data, col="slategrey",
     xlab="Year", 
     ylab="Patch reef area (m^2)", 
     main="Area versus year")
date.area.lm=lm(area_m2 ~ as.numeric(year), data=data)
abline(date.area.lm, col = "black")
summary(date.area.lm)
dev.copy(pdf,"output/date_area_effect.pdf")    
dev.off()
```

Now, let's see if there's anything going on with time when imagery resolution is considered:

```{r data analysis-1, echo=TRUE, results='hide'}

################################################
# analyze data
################################################

# Effects of imagery resolution + year

## Note: interpretation below is for my test data; your data's results may be different
data$year <- as.numeric(data$year) 
res.year.lm=lm(area_m2 ~ imagery_resolution_m * year, data=data) # run interaction model
summary(res.year.lm)      # overall model is significant, but not indiv parameters
anova(res.year.lm)        # run ANOVA to look at predictor variables in isolation
drop1(res.year.lm, test="F")  # use drop1 function to see which parameter can be dropped
res.year.lm2 <- lm(area_m2 ~ imagery_resolution_m + year, data=data)
                          # re-run model w/o interaction term that drop1 said to drop
summary(res.year.lm2)     # model now says both predictor variables are significant
drop1(res.year.lm2, test="F") # drop1 says neither remaining variable can be dropped

```

Or reef identity (reef number):

```{r data analysis-2, echo=TRUE, results='hide'}

################################################
# analyze data
################################################

# Effects of imagery resolution + reef ID
res.reefID.lm=lm(area_m2 ~ imagery_resolution_m * reef_no, data=data)
summary(res.reefID.lm)  
              # I get NAs for variable reef_no since only one reef used in my dataset
```

## Part IV: Discussion questions

* Discuss as a group any patterns that seemed to emerge...e.g.,
    + Which method you found to be more variable among groups/individuals - field or imagery?
    + How did imagery resolution affect your measurements?
    + What are the trade-offs between field vs imagery measurements?
    + What are the trade-offs between the different imagery a) platforms and b) resolutions?

* If time permits, move on to next section

## Part V: Write-up

* When you’re finished with your analyses, go back and give each figure a short caption 
* Structure your .Rmd file as a ‘mini-paper’ with a bit of very brief text (~2-5 sentences) for each of the usual paper sections: 
    + Introduction (summarize what you’re doing with this mini-project)
    + Methods (what you did, both in the field and the lab)
    + Results (here’s where your figures and stats outputs will go)
    + Discussion (mention any interesting findings and, if you have time, briefly relate this to any previous literature using a similar approach/asking similar questions)	
* This exercise will help you see the value of using R Markdown to write papers! 


